Mapping

### 2025-04-10 episodes/snps-phylogeny.Rmd
First touch up on SNPs phylogeny

In this episode we will try to pinpoint single nucleotide variants or single nucleotide polymorphism (SNPs) between our samples and the reference. The SNPs are determined by a process called read mapping in which they are aligned to the reference sequence.

@@

If you don't have all the trimmed reads, you can download them (will take a while, *dont't do it if you have the reads*):

@@

First we'll create a folder to hold the results from snippy:

```bash
interactive -A uppmax2025-3-4 -M snowy -t 04:00:00
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/data
mkdir snps
```

We need to load snippy:

```bash
module load bioinfo-tools snippy
```

Snippy is fast but a single run still takes about 15 minutes. We would therefore tell snippy to run all the samples after each other. However this time we cannot use a wildcard to do so. We would instead run all the samples in a loop. The "real" loop is commented out, and we use one that runs only the sample we've analyzed so far.

>>>>

In this episode we will try to pinpoint single nucleotide polymorphism (SNPs), also called single nucleotide variants (SNVs), between our samples and the reference. The SNPs are determined by a process called read mapping in which reads are aligned to the reference sequence.

@@

If you don't have all the trimmed reads, extract them from the archive in the `data` folder, using `tar`. `tar` is a program to take multiple files, including their folder structure and put them in a single `tar` "ball". Often the tarballs are compressed with `gzip`.

`tar` and `gzip` are computationally intensive and require you to start an interactive session.

@@

::: challenge

First we'll start the interactive session, load the snippy module and create a `snippy` folder in the `molepi/data` to hold the results from snippy.

::: solution

```bash
interactive -A uppmax2025-3-4 -M snowy -t 04:00:00
module load bioinfo-tools snippy
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/data
mkdir snps
```
::::::::::::

:::::::::::::
Snippy is fast but a single run still takes about 15 minutes. We would therefore tell snippy to run all the samples after each other. However this time we cannot use a wildcard to do so. We would instead run all the samples in a loop. The "real" loop is commented out, and we use one that runs only the sample we've analyzed so far.
